---
title: 'Exploring US COVID-19 Cases and Deaths - Johns Hopkins CSSE Data Sets'
author: "H. David Shea"
date: 2021-02-12
output:
    github_document: default
---

```{r setup, include = FALSE}
library(tidyverse)
library(rvest)
library(timetk)
library(lubridate)
library(broom)
library(knitr)
library(httr)

knitr::opts_chunk$set(
    connection = "secdb", # automatically uses this connection in sql chunks 
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)
```

## Johns Hopkins CSSE COVID-19 Data

### Field description

* State - The name of the State within the USA.
* Date - The most recent date the file was pushed.
* Lat - Latitude.
* Long_ - Longitude.
* Confirmed - Aggregated case count for the state.
* Deaths - Aggregated death toll for the state.
* Recovered - Aggregated Recovered case count for the state.
* Active - Aggregated confirmed cases that have not been resolved (Active cases = total cases - total recovered - total deaths).
* Incident_Rate - cases per 100,000 persons.
* Total_Test_Results - Total number of people who have been tested.
* Case_Fatality_Ratio - Number recorded deaths * 100/ Number confirmed cases.
* Testing_Rate - Total test results per 100,000 persons. The "total test results" are equal to "Total test results (Positive + Negative)" from COVID Tracking Project.

```{r rawdata, include = TRUE, warning = FALSE}
# source https://github.com/CSSEGISandData/COVID-19

csse_raw <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports_us"
csse_dir <- "https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports_us"
exclusions <- c("Diamond Princess", "Grand Princess" , "American Samoa", "Guam", "Northern Mariana Islands", "Recovered",
                "Virgin Islands", "Puerto Rico")
col_names <- c("State", "Country", "Date", "Lat", "Long_", "Confirmed", "Deaths", "Recovered", "Active", "FIPS", 
               "Incident_Rate", "Total_Test_Results", "People_Hospitalized", "Case_Fatality_Ratio", "UID", "ISO3", 
               "Testing_Rate", "Hospitalization_Rate")
col_classes <- c("character", "character", "Date", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric",
                 "numeric", "numeric", "numeric", "numeric", "numeric", "character",
                 "numeric", "numeric")

# This is unsatisfying, but it works - extract data file names from CSSE GitHub site
req <- GET(csse_dir)
stop_for_status(req)
fname <- tempfile(fileext = ".txt")
write_file(req$content, fname)
file_names <- tibble(text = readLines(fname)) %>%
  filter(str_detect(text,".csv")) %>%
  mutate(text = str_match(str_match(str_trim(text),"title=\"[1234567890-]*.csv\""),"[1234567890-]*.csv"))
unlink(fname)
rm(req, fname, csse_dir)

for(idx in seq_along(file_names$text)) {
  if(idx == 1) {
    us_states <- tibble(read.csv(paste(csse_raw, file_names[idx,], sep = "/"), col.names = col_names, colClasses = col_classes)) %>% 
      filter(!(State %in% exclusions)) %>% 
      select(-Country, -FIPS, -UID, -ISO3, -People_Hospitalized, -Hospitalization_Rate)
  } else {
    new_date <- tibble(read.csv(paste(csse_raw, file_names[idx,], sep = "/"), col.names = col_names, colClasses = col_classes)) %>% 
      filter(!(State %in% exclusions)) %>% 
      select(-Country, -FIPS, -UID, -ISO3, -People_Hospitalized, -Hospitalization_Rate)

    us_states <- rbind(us_states, new_date)
  }
}
rm(file_names, csse_raw, exclusions, col_names, col_classes, idx, new_date)

us_states[10:20,] %>% kable()
```

# Exploratory Data Analyses

```{r adjdata, include = TRUE, warning = FALSE}

# Create rolling average changes
us_states <- us_states %>% 
  mutate(State = as_factor(State)) %>%
  arrange(State, Date) %>% 
  group_by(State) %>% 
  mutate(
    Confirmed_7 = (Confirmed - lag(Confirmed, 7)) / 7,
    Deaths_7 = (Deaths - lag(Deaths, 7)) / 7,
    Recovered_7 = (Recovered - lag(Recovered, 7)) / 7,
    Active_7 = (Active - lag(Active, 7)) / 7,
    Incident_Rate_7 = (Incident_Rate - lag(Incident_Rate, 7)) / 7,
    Case_Fatality_Ratio_7 = (Deaths_7 *100) / Confirmed_7,
    Testing_Rate_7 = (Testing_Rate - lag(Testing_Rate, 7)) / 7
  )

#Create a national aggregate data set
us <- us_states %>%
  select(-Lat, -Long_) %>% 
  group_by(Date) %>%
  summarize(across(
    .cols = where(is.double),
    .fns = function(x) sum(x, na.rm = T),
    .names = "{col}"
  ))

save(us_states, file = "Data/jh_us_states.rdata")
save(us, file = "Data/jh_us.rdata")

us[10:20,] %>% kable()
```

# Basic Data Plots

```{r plots, include = TRUE, warning = FALSE}
coeff <- 30
g <- us %>%
  ggplot(aes(Date, Confirmed_7)) +
  geom_line(color = "orange") +
  theme(legend.position = "none") +
  geom_line(aes(x = Date, y = Deaths_7 * coeff), color = "red") +
  scale_y_continuous(
    labels = scales::comma,
    name = "Cases",
    sec.axis = sec_axis(Deaths_7 ~ . / coeff,
      name = "Deaths",
      labels = scales::comma
    )
  ) +
  theme(
    axis.title.y = element_text(color = "orange", size = 13),
    axis.title.y.right = element_text(color = "red", size = 13)
  ) +
  labs(
    title = "U.S. Cases vs. Deaths",
    subtitle = "7-Day Average",
    caption = "Source: JH CSSE",
    x = "Date"
  )
show(g)
```

