---
title: 'Exploring US COVID-19 Cases and Deaths - Johns Hopkins CSSE Data Sets'
author: "H. David Shea"
date: 2021-02-12
output:
    github_document: default
---

```{r setup, include = FALSE}
library(tidyverse)
library(rvest)
library(timetk)
library(lubridate)
library(broom)
library(knitr)
library(httr)

knitr::opts_chunk$set(
    connection = "secdb", # automatically uses this connection in sql chunks 
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)
```

## Johns Hopkins CSSE COVID-19 Data

## Dataset Overview

### Data Sets field descriptions

* State - The name of the State within the USA.
* Date - The most recent date the file was pushed.
* Lat - Latitude.
* Long_ - Longitude.
* Confirmed - Aggregated case count for the state.
* Deaths - Aggregated death toll for the state.
* Recovered - Aggregated Recovered case count for the state.
* Active - Aggregated confirmed cases that have not been resolved (Active cases = total cases - total recovered - total deaths).
* Incident_Rate - cases per 100,000 persons.
* Total_Test_Results - Total number of people who have been tested.
* Case_Fatality_Ratio - Number recorded deaths * 100/ Number confirmed cases.
* Testing_Rate - Total test results per 100,000 persons. The "total test results" are equal to "Total test results (Positive + Negative)" from COVID Tracking Project.

## Example Data

```{r rawdata, include = TRUE, warning = FALSE, echo = FALSE}
# source https://github.com/CSSEGISandData/COVID-19

# The r-script collect_JH_CSSE_Covid19_data.R is run daily at 4:00am via GitActions.  
# This process scrapes the JH CSSE site for data files and loads the US Data into data 
# frames and saves those to the Data directory.  The following lines load those data
# frames for this session.

load("Data/jh_us_states.rdata")
load("Data/jh_us.rdata")
```

### National Level Data

```{r us_sample, include = TRUE, warning = FALSE, echo = FALSE}
us[1:8,] %>% kable()
```

### State Level Data

```{r state_sample, include = TRUE, warning = FALSE, echo = FALSE}
us_states[1:8,] %>% kable()
```

## Exploratory Data Analyses

### US Cases versus Deaths

```{r us_cases_and_deaths, include = TRUE, warning = FALSE, echo = FALSE}
coeff <- 30
g <- us %>%
  ggplot(aes(Date, Confirmed_7)) +
  geom_line(color = "orange") +
  theme(legend.position = "none") +
  geom_line(aes(x = Date, y = Deaths_7 * coeff), color = "red") +
  scale_y_continuous(
    labels = scales::comma,
    name = "Cases",
    sec.axis = sec_axis(Deaths_7 ~ . / coeff,
      name = "Deaths",
      labels = scales::comma
    )
  ) +
  theme(
    axis.title.y = element_text(color = "orange", size = 13),
    axis.title.y.right = element_text(color = "red", size = 13)
  ) +
  labs(
    title = "U.S. Cases vs. Deaths",
    subtitle = "7-Day Average",
    caption = "Source:  Johns Hopkins University - Center for Systems Science and Engineering (CSSE)",
    x = "Date"
  )
show(g)
```

### US Fatality Ratio - Deaths as a Percent of Confirmed Cases

```{r us_fatality_ratio, include = TRUE, warning = FALSE, echo = FALSE}
us %>%
  ggplot(aes(Date, Case_Fatality_Ratio_7)) +
  geom_line() +
  geom_smooth() +
  labs(
    y = "7-Day Average Case Fatality Ratio",
    caption = "Source:  Johns Hopkins University - Center for Systems Science and Engineering (CSSE)",
    title = "Fatality Rate:  Deaths as a Percentage of Confirmed Cases"
  )
```

### US Testing Rate

```{r us_testing_rate, include = TRUE, warning = FALSE, echo = FALSE}
us %>%
  ggplot(aes(Date, Testing_Rate)) +
  geom_line() +
  geom_smooth() +
  labs(
    y = "Testing Rate",
    caption = "Source:  Johns Hopkins University - Center for Systems Science and Engineering (CSSE)",
    title = "U.S. Total Test Results per 100,000 Persons"
  )
```

### Selected States Cases versus Deaths

```{r states_cases_and_deaths, include = TRUE, warning = FALSE, echo = FALSE}
# ------------------------------------------
# select state by state analysis
state_subset <- c("New York", "Texas", "California", "Ohio", "Florida", "Massachusetts")

# illustrate selected states
coeff <- 30
g <- us_states %>%
  filter(State %in% state_subset) %>%
  ggplot(aes(Date, Confirmed_7)) +
  geom_line(color = "orange") +
  facet_wrap(~State, scales = "free") +
  theme(legend.position = "none") +
  geom_line(aes(y = Deaths_7 * coeff), color = "red") +
  scale_y_continuous(
    labels = scales::comma,
    name = "Cases",
    sec.axis = sec_axis(Deaths_7 ~ . / coeff,
      name = "Deaths",
      labels = scales::comma
    )
  ) +
  theme(
    axis.title.y = element_text(color = "orange", size = 13),
    axis.title.y.right = element_text(color = "red", size = 13)
  ) +
  labs(
    title = "Selected States Cases vs. Deaths",
    subtitle = "7-Day Average",
    caption = "Source:  Johns Hopkins University - Center for Systems Science and Engineering (CSSE)",
    x = "Date"
  )
show(g)
```

### Selected States Fatality Ratio - Deaths as a Percent of Confirmed Cases

```{r states_fatality_ratio, include = TRUE, warning = FALSE, echo = FALSE}
us_states %>%
  filter(State %in% state_subset) %>%
  ggplot(aes(Date, Case_Fatality_Ratio_7)) +
  geom_line() +
  geom_smooth() +
  facet_wrap(~State, scales = "free") +
  labs(
    y = "7-Day Average Case Fatality Ratio",
    caption = "Source:  Johns Hopkins University - Center for Systems Science and Engineering (CSSE)",
    title = "Fatality Rate:  Deaths as a Percentage of Confirmed Cases"
  )
```

### Selected States Testing Rate

```{r states_testing_rate, include = TRUE, warning = FALSE, echo = FALSE}
us_states %>%
  filter(State %in% state_subset) %>%
  ggplot(aes(Date, Testing_Rate)) +
  geom_line() +
  geom_smooth() +
  facet_wrap(~State, scales = "free") +
  labs(
    y = "Testing Rate",
    caption = "Source:  Johns Hopkins University - Center for Systems Science and Engineering (CSSE)",
    title = "Selected States Total Test Results per 100,000 Persons"
  )
```
