---
title:    'Exploring US COVID-19 Cases and Deaths'
subtitle: 'Our World in Data - Data Sets'
author:   'H. David Shea'
date: 2021-04-20
output:
    github_document: default
---

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(knitr)

knitr::opts_chunk$set(
    connection = "secdb", # automatically uses this connection in sql chunks 
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)

# various plots used in the analysis

cvd_plot <- function(data, title_lab) {
  coeff <- 30
  ggplot(data, aes(date, new_cases_smoothed)) +
    geom_line(color = "orange") +
    theme(legend.position = "none") +
    geom_line(aes(x = date, y = new_deaths_smoothed * coeff), color = "red") +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
    scale_y_continuous(
      labels = scales::comma,
      name = "Cases",
      sec.axis = sec_axis(new_deaths_smoothed ~ . / coeff,
                          name = "Deaths",
                          labels = scales::comma
      )
    ) +
    theme(
      axis.title.y = element_text(color = "orange", size = 13),
      axis.title.y.right = element_text(color = "red", size = 13)
    ) +
    labs(
      title = title_lab,
      subtitle = "7-Day Smoothed",
      caption = "Source:  Our World in Data",
      x = "Date"
    ) +
    scale_colour_brewer(palette = "Dark2")
}

cases_plot <- function(data, title_lab, color_lab) {
  ggplot(data, aes(date, new_cases_smoothed / 1000, color = location)) +
    geom_line() +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
    theme(legend.position = "bottom") +
    labs(
      title = title_lab,
      caption = "Source:  Our World in Data",
      color = color_lab,
      x = "Date",
      y = "7-Day Smoothed (000)"
    ) +
    scale_colour_brewer(palette = "Dark2")
}

deaths_plot <- function(data, title_lab, color_lab) {
  ggplot(data, aes(date, new_deaths_smoothed, color = location)) +
    geom_line() +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
    theme(legend.position = "bottom") +
    labs(
      title = title_lab,
      caption = "Source:  Our World in Data",
      color = color_lab,
      x = "Date",
      y = "7-Day Smoothed"
    ) +
    scale_colour_brewer(palette = "Dark2")
}

cases_facet_plot <- function(data, title_lab) {
  ggplot(data, aes(date, new_cases_smoothed / 1000)) +
    geom_line() +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
    facet_wrap(~location, scales = "free") +
    theme(legend.position = "bottom") +
    labs(
      title = title_lab,
      caption = "Source:  Our World in Data",
      x = "Date",
      y = "7-Day Smoothed Cases (000)"
    ) +
    scale_colour_brewer(palette = "Dark2")
}

deaths_facet_plot <- function(data, title_lab) {
  ggplot(data, aes(date, new_deaths_smoothed)) +
    geom_line() +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
    facet_wrap(~location, scales = "free") +
    theme(legend.position = "bottom") +
    labs(
      title = title_lab,
      caption = "Source:  Our World in Data",
      x = "Date",
      y = "7-Day Smoothed Deaths"
    ) +
    scale_colour_brewer(palette = "Dark2")
}

tests_plot <- function(data, title_lab, color_lab) {
  ggplot(data, aes(date, new_tests_smoothed_per_thousand, color = location)) +
    geom_line() +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %y") +
    theme(legend.position = "bottom") +
    labs(
      title = "New Tests per 1000 People",
      subtitle = title_lab,
      caption = "Source:  Our World in Data",
      color = color_lab,
      x = "Date",
      y = "7-Day Smoothed"
    ) +
    scale_colour_brewer(palette = "Dark2")
}

pr_plot <- function(data, title_lab, color_lab) {
  ggplot(data, aes(date, positive_rate, color = location)) +
    geom_line() +
    scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
    theme(legend.position = "bottom") +
    labs(
      title = "Positivity Rate",
      subtitle = title_lab,
      caption = "Source:  Our World in Data",
      color = color_lab,
      x = "Date",
      y = "7-Day Rolling Average"
    ) +
    scale_colour_brewer(palette = "Dark2")
}

```

## Our World in Data (OWID) COVID-19 Data

This analysis uses the Our World in Data COVID-19 data sets.  These are available at the [_OWID GitHub site_](https://covid.ourworldindata.org/).

```{r rawdata, include = TRUE, warning = FALSE, echo = FALSE}
# source https://covid.ourworldindata.org/

# OWID Corona Virus data
owid_full_metadata_url <- "https://covid.ourworldindata.org/data/owid-covid-codebook.csv"
owid_full_dataset_url <- "https://covid.ourworldindata.org/data/owid-covid-data.csv"

# Get metadata
owid_full_metadata <- tibble(read.csv(owid_full_metadata_url))

# Get the actual data
owid_full_dataset <- tibble(read.csv(owid_full_dataset_url)) %>%
  mutate(date = as.Date(date)) %>%
  arrange(location, date)

# location groupings used in analyses
top_50_pop <- owid_full_dataset %>%
  filter(date == (max(unique(owid_full_dataset$date)) - 1)) %>%
  filter(!str_starts(iso_code, "OWID")) %>%
  arrange(desc(population)) %>%
  filter(row_number() <= 50) %>%
  pull(location)

regions <- c("Africa", "Asia", "Europe", "North America", "Oceania", "South America")
G7 <- c("Canada", "France", "Germany", "Italy", "Japan", "United Kingdom", "United States")

# select countries (from top 50 populations) by region
americas <- c("United States", "Brazil", "Mexico", "Colombia", "Argentina", "Canada", "Peru", "Venezuela")
se_asia <- c("India", "Indonesia", "Pakistan", "Bangladesh", "Philippines", "Vietnam", "Thailand", "Myanmar", "Malaysia")
west_asia <- c("China", "Japan", "South Korea")
mid_east <- c("Iran", "Iraq", "Saudi Arabia")
africa <- c("Nigeria", "Ethiopia", "Egypt", "Tanzania", "South Africa", "Kenya", "Uganda", "Algeria", "Sudan")
europe <- c("Germany", "France", "United Kingdom", "Italy", "Spain", "Russia", "Turkey", "Ukraine", "Poland")

```

## Dataset Overview

### Data Sets field descriptions

```{r metadata, include = TRUE, warning = FALSE, echo = FALSE}
owid_full_metadata %>% select(column, description) %>% kable()
```

## Example Data

```{r dataset, include = TRUE, warning = FALSE, echo = FALSE}
(owid_full_dataset %>% arrange(date, location))[401:410,] %>% kable()
```

## Exploratory Data Analyses

### World Cases versus Deaths

```{r global_cases_and_deaths, include = TRUE, warning = FALSE, echo = FALSE}
owid_full_dataset %>%
  filter(location == "World") %>% 
  cvd_plot("World Cases vs. Deaths")
```

### US Cases versus Deaths

```{r us_cases_and_deaths, include = TRUE, warning = FALSE, echo = FALSE}
owid_full_dataset %>%
  filter(location == "United States") %>% 
  cvd_plot("U.S. Cases vs. Deaths")
```

### Global Regions

```{r region_cases_and_deaths, include = TRUE, warning = FALSE, echo = FALSE}
owid_full_dataset %>%
  filter(location %in% regions) %>% 
  cases_plot("Cases by Region", "Region")

owid_full_dataset %>%
  filter(location %in% regions) %>% 
  deaths_plot("Deaths by Region", "Region")
```

### Current Problem Areas
```{r problem_areas, include = TRUE, warning = FALSE, echo = FALSE}
top_6_change_cases <- owid_full_dataset %>%
  filter(date >= (max(unique(owid_full_dataset$date)) - 35)) %>%
  filter(location %in% top_50_pop) %>%
  mutate(
    cases_change = round(total_cases / lag(total_cases, 30), 2),
    deaths_change = round(total_deaths / lag(total_deaths, 30),2)
    ) %>%
  filter(date == (max(unique(owid_full_dataset$date)) - 1)) %>%
  select(location, date, cases_change) %>%
  arrange(desc(cases_change)) %>%
  filter(row_number() <= 6)

top_6_change_deaths <- owid_full_dataset %>%
  filter(date >= (max(unique(owid_full_dataset$date)) - 35)) %>%
  filter(location %in% top_50_pop) %>%
  mutate(
    cases_change = total_cases / lag(total_cases, 30),
    deaths_change = total_deaths / lag(total_deaths, 30)
  ) %>%
  filter(date == (max(unique(owid_full_dataset$date)) - 1)) %>%
  select(location, date, deaths_change) %>%
  arrange(desc(deaths_change)) %>%
  filter(row_number() <= 6)
```

#### Countries with Highest Change in Cases Over the Past Month

```{r problem_areas_cases, include = TRUE, warning = FALSE, echo = FALSE, out.width = "90%"}
top_6_change_cases %>% kable()

owid_full_dataset %>%
  filter(location %in% (top_6_change_cases %>% pull(location))) %>%
  cases_facet_plot("Cases in Country")
```

#### Countries with Highest Change in Deaths Over the Past Month

```{r problem_areas_deaths, include = TRUE, warning = FALSE, echo = FALSE, out.width = "90%"}
top_6_change_deaths %>% kable()

owid_full_dataset %>%
  filter(location %in% (top_6_change_deaths %>% pull(location))) %>%
  deaths_facet_plot("Deaths in Country")
```

### Testing

#### Tests per 1000 People

```{r test_per_1000, include = TRUE, warning = FALSE, echo = FALSE}
filter_and_plot <- function(filter, label) {
  # Most countries were reporting tests by May 2020
  xlim <- ymd('2020-05-01')
  
  owid_full_dataset %>%
    filter(location %in% filter) %>% 
    filter(date >= xlim) %>% 
    arrange(location, date) %>% 
    select(location, date, new_tests_smoothed_per_thousand) %>% 
    tests_plot(label, "Country")
}

filter_set <- list(G7, americas, se_asia, west_asia, mid_east, africa, europe)

label_set <- list("G7 Countries", "Americas (Larger Populations)", "Southeast Asia (Larger Populations)", 
               "Western Asia (Larger Populations)", "Mideast (Larger Populations)", "Africa (Larger Populations)", 
               "Europe (Larger Populations)")

for(idx in seq_along(filter_set)) {
  show(filter_and_plot(c(filter_set[[idx]]), label_set[[idx]]))
}

rm(filter_and_plot, filter_set, label_set)
```

#### Positivity Rates

```{r positivity_rate, include = TRUE, warning = FALSE, echo = FALSE}
filter_and_plot <- function(filter, label) {
  # Most countries were reporting tests by May 2020
  xlim <- ymd('2020-05-01')
  
  owid_full_dataset %>%
    filter(location %in% filter) %>% 
    filter(date >= xlim) %>% 
    arrange(location, date) %>% 
    select(location, date, positive_rate) %>% 
    pr_plot(label, "Country")
}

filter_set <- list(G7, americas, se_asia, west_asia, mid_east, africa, europe)

label_set <- list("G7 Countries", "Americas (Larger Populations)", "Southeast Asia (Larger Populations)", 
               "Western Asia (Larger Populations)", "Mideast (Larger Populations)", "Africa (Larger Populations)", 
               "Europe (Larger Populations)")

for(idx in seq_along(filter_set)) {
  show(filter_and_plot(c(filter_set[[idx]]), label_set[[idx]]))
}

rm(filter_and_plot, filter_set, label_set)
```
